p-adic stability in linear algebra

### 2015-01-18 examples/stability.tex
Small fixes

The notation $M_{m,n}(K)$ refers to the $K$-vector space of matrices 
with $m$ rows and $n$ columns with coefficients in $K$.  We will repeatedly
use the Smith decomposition for $M \in M_{m,n}(K)$:
\[
M = U_M \cdot \Delta_M \cdot V_M,
\]
with $U_M$ and $V_M$ unimodular and $\Delta_M$ diagonal.  Write $\sigma_i(M)$
for the valuation of the $(i,i)$-th entry of $\Delta_M$, and by convention set
$\sigma_i(M) = +\infty$ if $i > \min(m,n)$.  Order the $\sigma_i(M)$ so that $\sigma_i(M) \le \sigma_{i+1}(M)$.

@@

\begin{proof}
We write $A \cdot dB + dA \cdot B = U_A \cdot M \cdot V_B$ with
$$M = \Delta_A \cdot V_A \cdot dA \cdot V_B^{-1} 
+ U_A^{-1} \cdot dB \cdot U_B \cdot \Delta_B.$$
When $dA$ varies in $M_{a,b}(\O_K)$ so does $V_A \cdot dA \cdot V_B^{-1}$
and therefore the first summand in $M$ varies in the subspace of 
$M_{a,c}(\O_K)$ consisting of matrices whose all entries on the $i$-th
row have valuation at least $a_i$. Arguing similarly for the second
summand, we deduce the first statement of the Proposition. The second
statement is left to the reader.
\end{proof}

Let us briefly comment on Proposition \ref{prop:mulmatrix} from the 
perspective of precision. The second statement roughly means that when 
we are computing the product $AB$, we are gaining $\sum_{m=1}^\infty 
a^{(m)} b^{(m)}$ significant digits in absolute precision\footnote{We 
note that, on the other side, the valuation of the entries of $AB$ may 
increase, meaning that we are also loosing some significant digits if we 
are reasoning in relative precision.} as soon as $N > \min(a_r, b_t)$ 
(\emph{cf} Proposition \ref{prop:precision2}). However these digits are 
not --- even individually --- attached to some particular entry of $AB$ 
but are ``spread out'' among all its entries. Thus, a usual 
coordinate-wise precision data does not see this gain of precision in 
general. There is a more pedestrian way to visualize this phenomenon. 
Remark that the product $AB$ can be rewritten:
$AB = U_A \cdot P \cdot V_B$ with
$P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$.
Tracking precision in the usual way, we see that the $(i,j)$-th entry
of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$. Hence, the 
gain of precision is visible here. However it is then ``spread out'' 
when we are computing $U_A \cdot P \cdot V_B$ and mostly disappears in 
the usual coordinate-wise precision model. The lattice precision model 
then appears as a way to keep track of this ``diffused'' gain of
precision.

>>>>

The notation $M_{m,n}(K)$ refers to the $K$-vector space of matrices 
with $m$ rows and $n$ columns with coefficients in $K$.  We will repeatedly
use the Smith decomposition for $M \in M_{m,n}(K)$:
\[
M = U_M \cdot \Delta_M \cdot V_M,
\]
with $U_M$ and $V_M$ unimodular and $\Delta_M$ diagonal.  Write $\sigma_i(M)$
for the valuation of the $(i,i)$th entry of $\Delta_M$, and by convention set
$\sigma_i(M) = +\infty$ if $i > \min(m,n)$.  Order the $\sigma_i(M)$ so that $\sigma_i(M) \le \sigma_{i+1}(M)$.

@@

\begin{proof}
We write $A \cdot dB + dA \cdot B = U_A \cdot M \cdot V_B$ with
$$M = \Delta_A \cdot V_A \cdot dA \cdot V_B^{-1} 
+ U_A^{-1} \cdot dB \cdot U_B \cdot \Delta_B.$$
When $dA$ varies in $M_{a,b}(\O_K)$ so does $V_A \cdot dA \cdot V_B^{-1}$
and therefore the first summand in $M$ varies in the subspace of 
$M_{a,c}(\O_K)$ consisting of matrices whose all entries on the $i$th
row have valuation at least $a_i$. Arguing similarly for the second
summand, we deduce the first statement of the Proposition. The second
statement is left to the reader.
\end{proof}

Let us briefly comment on Proposition \ref{prop:mulmatrix} from the 
perspective of precision. The second statement roughly means that when 
we are computing the product $AB$, we are gaining $\sum_{m=1}^\infty 
a^{(m)} b^{(m)}$ significant digits in absolute precision\footnote{We 
note that, on the other side, the valuation of the entries of $AB$ may 
increase, meaning that we are also loosing some significant digits if we 
are reasoning in relative precision.} as soon as $N > \min(a_r, b_t)$ 
(\emph{cf} Proposition \ref{prop:precision2}). However these digits are 
not --- even individually --- attached to some particular entry of $AB$ 
but are ``spread out'' among all its entries. Thus, a usual 
coordinate-wise precision data does not see this gain of precision in 
general. There is a more pedestrian way to visualize this phenomenon. 
Remark that the product $AB$ can be rewritten
$AB = U_A \cdot P \cdot V_B$ with
$P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$.
Tracking precision in the usual way, we see that the $(i,j)$th entry
of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$. Hence, the 
gain of precision is visible here. However it is then ``spread out'' 
when we are computing $U_A \cdot P \cdot V_B$ and mostly disappears in 
the usual coordinate-wise precision model. The lattice precision model 
then appears as a way to keep track of this ``diffused'' gain of
precision.
